Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-08-22 Multiresolution_framework/ltx/response_to_reviews.tex

        \item The authors say that a multi-resolution approach would explain the``organizational level of the brain at which cognition and behaviour are best explained'' but it is unclear what the alternative levels considered here are and what the significance of the results obtained is.

\emph{Within our framework, the alternate levels are defined by the spatial scales of the wavelet decomposition. The significant result is the formulation of a model when various spatial scales of a neural field can be simultaneously represented. The ability to image the multi-dimensional neural field will lead to a better understanding of the important spatial scales when neural information processing occurs.}

		\item Similar claims to the ones mentioned here regarding implications for ``patient-specific'' modeling have also been made in Freestone et al. (2011), so it is not clear how these relate to the current multi-resolution approach and also such remarks might look too ambitious given that only synthetic data with fixed parameters are considered.

	\emph{Please refer to Response 2 to the handling editor. In summary, the framework can be used to create patient-specific models when it is applied with patient-specific data. Demonstrating convergence with synthetic data, where actual states and parameters are known, is a necessary first step. Furthermore, there is a significant body of work where various phenomena have been replicated using the neural field equations. This motivates the creation of tools to verify the models and experimentally test hypotheses that computational studies have generated.}

@@

 % 
 % \begin{figure*}[!ht]
 % \begin{center}
 % \includegraphics{./Graph/pdf/fig1.pdf} 
 % \end{center}
 % \caption{}
 % \label{fig:Figure1}
 % \end{figure*}

\end{enumerate}

    \section{Reviewer 3}  
    \subsection{Main Concerns}
		\begin{enumerate} 
			\item I am not at all sure that isotopic scaling of the multi-resolution is appropriate for brain connections, but I cannot say at this point that it is not.

			\emph{It is generally accepted that the intracortical connectivity structure is isotropic. However, as the reviewer pointed out, this is not the case for cortico-cortical connectivity. Nevertheless, the presented framework is also valid for other connectivity kernels provided they are spatially homogeneous and that they can be approximated by a set of weighted isotropic basis functions (that can form a net anisotropic kernel). The connectivity kernel can take any arbitrary shapes but this may require more basis functions for an accurate reconstruction.}

@@

			\item This model cries out for validation in real neuronal systems, but alas, at this point we only see convergence to synthetic data.

			 \emph{We agree with this statement 100 percent. We hope that this paper will motivate new experiments, where neural field models can be experimentally validated. It has now been nearly 40 years since neural field models where introduced in computational studies. For the first time we are at the exciting point in time where computing power and estimation algorithms are sufficiently advanced to link models to actual data. This is why we believe the framework the is introduced in this paper has the potential to make an impact. For further remarks regarding this point please refer to Response 2 to the handling editor.}
		 % 			
		 % \emph{\parham{The authors appreciate that the only way to truly validate the proposed framework is with real data. However, before validating the method on real data, we believe that it is essential to first validate the estimation method using accepted models and synthetic data, where the true and estimated parameters can be compared. We believe this was achieved in this paper. The title of the manuscript was chosen to reflect this fact, stating this is a multi-resolution framework for the Amari type neural field model. In an effort to make the method employed in the paper more explicit, we have added ``data was generated synthetically using Amari neural field model'' to the Results section.}}

@@

		\item This audience needs to have all symbols, including commonly used ones such as `equality over distribution', double font domains like R and Z, and jargon terms such as `lead field'.

		\emph{We have defined independent and identically distributed (i.i.d.) random variables under Eq. (5). The double font domains are added to Table 1. A brief description of `lead field' is also added below Eq. (5).}

		\item Frankly, I would off-load all of the very difficult mathematics to the appendices and supplementary data, and focus a NeuroImage paper on the results of image analysis.

			\emph{We attempted to move the derivation from Eq. 22 to Eq. 33 to the appendix. This led to more complications, where it was very difficult to define the various components of the model and a notational nightmare. To guide the reader who may just be interested in the conceptual framework and results we have added ``The conceptual framework and the key results can be interpreted without following this derivation in detail.'' below Eq. 22.}

@@

 			\emph{\parham{We have added a footnote on page 17 : The Python source code for our numerical simulations is available from the authors on request.}}

\dean{Cant we just upload it?}

			\end{enumerate}  
			\subsection{Moderate Concerns} 
			\begin{enumerate} 
			 \item Page 2: the authors should clearly define the difference between neural field and neural mass models.

			\emph{We have added the following paragraph on page 2: The main difference between these two classes is that neural field models represent a cortical sheet, making a continuum approximation to describe the spatiotemporal evolution of neural dynamics (such as mean firing rates or mean membrane potentials), while neural mass models describe ensembles of neurons at a localized point, which may be networked to multiple points. Both neural field and neural mass models make a mean-field approximation to describe the mass action of the brain.}

### 2012-08-22 Multiresolution_framework/ltx/manuscript.tex
edits to discussion

The utility of these models for answering questions in clinical neuroscience and neurology lies in the ability to infer physiological parameters from electrophysiological data. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. Large-scale neuroinformatic efforts, such as the Virtual Brain Project (www.thevirtualbrain.org \citep{Jirsa2010}), are on the way to enable patient-specific full brain modeling. This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. \dean{The main difference between these two classes is that neural field models represent a cortical sheet, making a continuum approximation to describe the spatiotemporal evolution of neural dynamics (such as mean firing rates or mean membrane potentials), while neural mass models describe ensembles of neurons at a localized point, which may be networked to multiple points. Both neural field and neural mass models make a mean-field approximation to describe the mass action of the brain.} It is expected that parameters of these models will be patient-specific and will, therefore, need to be inferred from data.

@@

The aim of the next part of the derivation is to isolate the state vector on the left hand side of the equation. To do this we first multiply both sides of Eq.~\eqref{eq:ApproxDiscreteTimeModel} by $\boldsymbol\mu\left(r\right)$ and integrate over space giving 

@@

\parham{\subsection{State-Space Representation}
In order to apply estimation algorithms to neural field models a finite-dimensional state-space model must be formed. The problem of representing a continuous field with a finite number of states is solved by the basis function decomposition. An alternative to using the B-spline wavelets is Gaussian basis functions, which was adopted by \cite{Freestone2011} to decompose the field and connectivity kernel. One clear advantage of B-spline scaling and wavelet functions is the MRA capability, allowing the simultaneous representation of the neural field at different spatial scales. B-spline scaling and wavelet functions are compactly supported, eliminating  numerical errors caused by the Gaussian basis function. The partition of unity of these basis functions enable the reconstruction of the neural field when it saturates, this may be beneficial to model neurological disorders such as epilepsy. Attaining fine scale approximation of the neural field using Gaussian basis functions results into an ill-posed problem as the number of states is increased significantly, however, this is not the case when using B-spline scaling and wavelet functions due to their semi-orthogonality property.\\ \cite{Freestone2011} showed experimentally that Gaussian functions are a suitable choice for modeling the observation kernel when intracranial measurements are considered. B-splines converge to Gaussian functions as the order of the B-spline, $m$, tends to infinity. The cubic B-spline is already a very close approximation to the Gaussian function, therefore these should also be suitable for modeling the observation kernel while having the advantage of being compactly supported.}

\subsection{Estimation Algorithm}
\parham{The implementation of the M-step in this work significantly simplifies the computational complexity of the EM based algorithm proposed by \cite{Dewar2009} as it is shown in Table~\ref{table:MRA-ComputationalComplexity}, namely from $O(n_x^4n_{\theta}^2)$ to $O(n_x^2n_{\theta}^2)$. This is crucial for high dimensional systems and more specifically for implementing the multi-resolution framework introduced herein which otherwise would be possible.}

@@

\parham{The basis structure (i.e., non-redundant, versus a (tight) frame representation) is a requirement for this framework.                                              }Choosing isotropic scaling and wavelet functions allows switching the order of operations with the inner product and convolution in Eq.~\eqref{eq:ApproxDiscreteTimeModel4}. This simplifies the computational complexity of the estimation algorithm significantly. Although this is correct for one dimensional B-spline scaling and wavelet functions, it does not hold at higher dimensions as the resulting B-spline wavelets are not isotropic. One alternative could be  
isotropic polyharmonic B-splines scaling functions and wavelets \citep{VanDeVille2005}.

 \subsection{Modification to Higher Dimensions}
 The reconstruction of the neural field in higher dimensions requires $2^n-1$ wavelets, where $n$ is the dimension of the space. For example a two-dimensional MRA can be implemented using 2-D scaling and wavelet functions built up using the tensor-product approach \citep{Meyer1992}. In this case three wavelet functions are required to  extract fine features of the field at vertical, horizontal and diagonal orientations. The formulations provided in this paper can be easily extended to two dimensions. However, the computational load of the estimation algorithm will be significant. It is important to investigate methods for choosing basis functions to balance the complexity with the reconstruction capability of the model. \parham{It is also possible to apply other than dyadic subsampling schemes in multiple dimensions, which leads to less wavelet subbands. For example, in 2D, the quincunx scheme is well-known to have better isotropic properties. Unfortunately, in 3D, it is not possible to have a subsampling operator (acting on a Cartesian sampling lattice) that is performing subsampling at equal rate in each dimension at each iteration \citep{VanDeVille2005a}.}
\subsection{Extensions to the Framework} 
Clearly there is, and will always be, a discrepancy between the dynamics of the model and the experimental observation. Nevertheless, the model-based framework proposed in this paper may enable meaningful state tracking and connectivity estimation. The MRAIDE approach is not limited to neural fields; the framework can be applied to modeling other multi-resolution spatiotemporal dynamical systems such as weather systems, ecological systems, and others~\citep{Wikle2002,Xu2005}. The key development is the multi-resolution decomposition forming the state-space model. While the decomposition holds for more sophisticated models, efficiently performing nonlinear smoothing in the large state-space remains a challenge.

>>>>

The utility of these models for answering questions in clinical neuroscience and neurology lies in the ability to infer physiological parameters from electrophysiological data. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. Large-scale neuroinformatic efforts, such as the Virtual Brain Project (www.thevirtualbrain.org \citep{Jirsa2010}), are on the way to enable patient-specific full brain modeling. This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. The main difference between these two classes is that neural field models represent a cortical sheet, making a continuum approximation to describe the spatiotemporal evolution of neural dynamics (such as mean firing rates or mean membrane potentials), while neural mass models describe ensembles of neurons at a localized point, which may be networked to multiple points. Both neural field and neural mass models make a mean-field approximation to describe the mass action of the brain. It is expected that parameters of these models will be patient-specific and will, therefore, need to be inferred from data.

@@

The aim of the next part of the derivation is to isolate the state vector on the left hand side of the equation. The conceptual framework and the key results can be interpreted without following this derivation in detail. To isolate the state vector we first multiply both sides of Eq.~\eqref{eq:ApproxDiscreteTimeModel} by $\boldsymbol\mu\left(r\right)$ and integrate over space giving 

@@

\subsection{State-Space Representation}
In order to apply estimation algorithms to neural field models a finite-dimensional state-space model must be formed. The problem of representing a continuous field with a finite number of states is solved by the basis function decomposition. An alternative to using the B-spline wavelets are Gaussian basis functions, which were adopted by \cite{Freestone2011} to decompose the field and connectivity kernel. One clear advantage of B-spline scaling and wavelet functions is the MRA capability, allowing the simultaneous representation of the neural field at different spatial scales. B-spline scaling and wavelet functions are compactly supported, eliminating  numerical errors caused by the Gaussian basis function. The partition of unity (ability to represent a zero spatial frequency field) of these basis functions enable the reconstruction of the neural field when it saturates. This may be beneficial to model neurological disorders such as epilepsy. Attaining fine-scale approximation of the neural field using Gaussian basis functions results in an ill-posed problem as the number of states is increased significantly. However, this is not the case when using B-spline scaling and wavelet functions due to their semi-orthogonality property.\\ \citet{Freestone2011} showed experimentally that Gaussian functions are a suitable choice for modeling the observation kernel when intracranial measurements are considered. B-splines converge to Gaussian functions as the order of the B-spline, $m$, tends to infinity. The cubic B-spline is already a very close approximation to the Gaussian function. Therefore, these should also be suitable for modeling the observation kernel while having the advantage of being compactly supported.

\subsection{Estimation Algorithm}
The implementation of the M-step in this work significantly simplifies the computational complexity of the EM based algorithm proposed by \cite{Dewar2009}, as shown in Table~\ref{table:MRA-ComputationalComplexity} ($O(n_x^4n_{\theta}^2)$ to $O(n_x^2n_{\theta}^2)$). This is crucial for high-dimensional systems, and more specifically, for implementing the multi-resolution framework introduced herein. Furthermore, choosing isotropic scaling and wavelet functions allows switching the order of operations with the spatial convolution and inner product in Eq.~\eqref{eq:ApproxDiscreteTimeModel4}. This simplifies the computational complexity of the estimation algorithm significantly.

@@

\dean{The basis structure (i.e., non-redundant versus a (tight) frame representation) is a requirement for this framework.}

\subsection{Modification to Higher Dimensions}
The reconstruction of the neural field in higher dimensions requires $2^n-1$ wavelets, where $n$ is the dimension of the space. For example a two-dimensional MRA can be implemented using 2-D scaling and wavelet functions built up using the tensor-product approach \citep{Meyer1992}. In this case three wavelet functions are required to extract fine features of the field at vertical, horizontal and diagonal orientations. Theoretically, the formulations provided in this paper can be easily extended to two dimensions. However, the increase in computational load of the estimation algorithm will be significant. This is not only due to the increase in number of wavelets, but also the switch of convolution and inner product does not hold, as the resulting higher-dimensional B-spline wavelets are not isotropic. A possible solution could be to use  
isotropic polyharmonic B-splines scaling functions and wavelets as an alternative \citep{VanDeVille2005}.

It is important to investigate methods for choosing basis functions to balance the complexity with the reconstruction capability of the model. It is also possible to apply other methods than dyadic subsampling schemes in multiple dimensions, which leads to less wavelet subbands. For example, in 2D, the quincunx scheme is well-known to have better isotropic properties. Unfortunately, in 3D, it is not possible to have a subsampling operator (acting on a Cartesian sampling lattice) that is performing subsampling at equal rate in each dimension at each iteration \citep{VanDeVille2005a}.
\subsection{Extensions to the Framework} 
Clearly there is, and will always be, a discrepancy between the dynamics of the model and the experimental observation. Nevertheless, the model-based framework proposed in this paper may enable meaningful state tracking and connectivity estimation. The MRAIDE approach is not limited to neural fields; the framework can be applied to modeling other multi-resolution spatiotemporal dynamical systems such as weather systems, ecological systems, and others~\citep{Wikle2002,Xu2005}. The key development is the multi-resolution decomposition forming the state-space model. While the decomposition holds for more sophisticated models, efficiently performing nonlinear smoothing in the large state-space remains a challenge.
